Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-02 latex/main.tex
Updates from Overleaf

As new data became available during the pandemic,
volunteers continually detected recombinants which might have concerning features,
particularly increased transmissibility, disease severity, and
ability to escape natural or vaccine-induced immunity (e.g., \cite{Tamura2022}).
The process of gathering evidence for and characterizing such recombinants is manual,
involving visual inspection of phylogenies for unusual branches
caused by forcibly placing recombinant sequences onto them,
typically followed by visual analysis of genome sequence alignments
to identify parental lineages and breakpoint locations.
This process has resulted in the designation of recombinant Pango lineages
(whose aliases begin with an “X”), to which we refer as “Pango Xs”.
Below, we summarize our analysis of the Pango Xs in the ARG
(for extended analyses, see Document S1 and Document S2).

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.85\linewidth]{pango_x_subgraphs.png}
    \caption{Illustrative subgraphs in the ARG. Subgraphs around the XA-associated recombination node (depicted as an encircled R) (A); the recombination node associated with a group of “nested” Pango Xs (B); and the MRCA of the putative non-recombinants XN and XAU (C). For more detailed visualizations of these subgraphs, see Document S2.}
    \label{fig:subgraphs}
\end{figure}

\subsection*{Table 1. Recombination events associated with Pango Xs.}

@@

%%% TODO: Mention RIPPLES somewhere here? Or just in Discussion?
We compared sc2ts against three state-of-the-art (non-phylogenetic) methods that detect inter-lineage recombinants
by scanning sequences for mutations associated with different Pango lineages:
(1) RecombinHunt \cite{Alfonsi2024},
(2) CovRecomb \cite{Li2024}, and
(3) rebar (https://github.com/phac-nml/rebar).

@@

\begin{tabular}{|l | l | l | l|}
 \hline
 \textbf{Method} & \textbf{Concordant in parental lineages (\%)} & \textbf{Concordant in breakpoint interval (\%)} & \textbf{Distance (bp)} \\ [1ex]
 \hline
 Sc2ts & XX & XX & XX \\ [1ex]
 \hline
 RecombinHunt-GISAID & XX & XX & XX \\ [1ex]
 \hline
 RecombinHunt-Nextstrain & XX & XX & XX \\ [1ex]
 \hline
 CovRecomb & XX & XX & XX \\ [1ex]
 \hline
 Rebar & XX & XX & XX \\ [1ex]
 \hline
\end{tabular}

@@

We find an overall preponderance of breakpoint intervals towards the 3’ end of the genome,
particularly at the left boundary of the Spike gene (\ref{fig:landscape}),
consistent with the pattern of excess observed in previous studies \cite{Turakhia2022,Li2024}.
While this might result from a higher recombination rate in this region,
it may also be due to the higher levels of polymorphism in the Spike gene,
allowing recombination endpoints to be estimated more precisely.
Reassuringly, sites having unusually high mutation counts are absent in this region (\ref{fig:landscape}),
which would confound recombination detection.

Furthermore, we find a negative relationship between
breakpoint interval size and the amount of divergence between recombinant parents,
which is measured by time to their MRCA (\ref{fig:landscape}).
When the two parental lineages have diverged for longer periods of time (the rightmost points),
more information becomes available on either side of a breakpoint to pinpoint its location, resulting in shorter intervals.
For instance, the breakpoint intervals tend to be the smallest for events involving recombination
between the deeply divergent Omicron BA.1 and BA.2 sublineages (the red points in \ref{fig:landscape}).

%%% TODO: Old version as placeholder. This figure is not yet finalised.
\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{recombinant_intervals}
    \caption{Recombination intervals. (A) Distribution of the breakpoint
intervals of the 356 ``robust'' recombination events along the genome
(annotated with gene labels on top). Each line represents an interval colored by its size.
(B) Number of overlapping breakpoint intervals.
(C) FIXME Heatmap showing the co-occurrence counts of the parents classified by Scorpio designations.
(D) Relationship between breakpoint interval size and time between the recombination event
and the parental MRCA. The black curve shows the theoretical expected relationship
between these quantities, estimated by minimizing the sum of squared residuals
given a single parameter (the mutation rate) (STAR Methods).}
    \label{fig:landscape}
\end{figure}

@@

We contribute several novel methods and concepts which may find applications beyond sc2ts.
First, we devise a simplified LS model, which extends the classic concept of parsimony.
In sc2ts, this model allows for accelerated HMM likelihood calculations,
enabling reconstruction of the pandemic-scale ARG presented here.
Existing HMMs used in statistical genetics may be modified in a similar fashion for significant speedup.
Second, we propose the concept of mutations averted by recombination,
which leads to an interpretable quantity to weigh
the support for the hypothesis that a given sample is a recombinant
against the hypothesis that it is a non-recombinant,
considering the sampled genetic diversity in an ARG and a specific HMM.
Third, we propose the concept of HMM cost to identify poorly explained samples
which are either too divergent or have a high number of mutations inconsistent with their collection dates.
Here we use HMM cost to reject time traveling samples,
which no existing phylogenetic method detects and handles during inference
(e.g., in the Viridian UShER phylogeny, parent and child nodes do not necessarily follow time order \cite{Hunt2024}),
resulting in a clean dataset rid of samples with erroneous collection dates that may affect downstream analyses.
Future work may be estimating dates for the suspected time traveling samples using the ARG provided here
and the sc2ts sample matching engine.

Recombination is an important evolutionary force in viruses \cite{Perez-Losada2015},
but has been understudied in part due to a lack of suitable data models, methods, and tools \cite{Neches2020}.
While the methods and tools developed here are tailored for SARS-CoV-2,
they can be extended and adapted for other viruses.
A major advantage of the sc2ts methods and tools is
that they are built on top of the mature software libraries of tskit and VCF-Zarr,
which enable pandemic-scale analyses that can be performed on typical laptops.
Both tskit and VCF-Zarr are developed in Python and leverage the popular PyData ecosystem,
making them highly adaptable for a broad range of applications.
Together with the pandemic-scale ARG, we hope that sc2ts will be highly accessible to many researchers,
enabling and accelerating studies of SARS-CoV-2 for years to come.

>>>>

As new data became available during the pandemic,
volunteers continually detected recombinants which might have concerning features,
particularly increased transmissibility, disease severity, and
ability to escape natural or vaccine-induced immunity (e.g., XBB\cite{Tamura2022}).
The process of gathering evidence for and characterizing such recombinants is manual,
involving visual inspection of phylogenies for unusual branches
caused by forcibly placing recombinant sequences onto them,
typically followed by visual analysis of genome sequence alignments
to identify parental lineages and breakpoint locations.
This process has resulted in the designation of recombinant Pango lineages
(whose aliases begin with an “X”), to which we refer as “Pango X”.
Below, we summarize our analysis of the Pango Xs in the ARG
(for extended analyses, see Document S1 and Document S2).

@@

\begin{figure}[h]
\centering
\includegraphics[width=0.85\linewidth]{pango_x_subgraphs.png}
\caption{Illustrative subgraphs in the ARG. Subgraphs around the XA-associated recombination node (A); the recombination node associated with a group of “nested” Pango Xs (B); and the MRCA of the non-recombinants XN and XAU (C). For more detailed visualizations of these subgraphs, see Document S2.}
\label{fig:subgraphs}
\end{figure}

@@

%%% TODO: Mention RIPPLES somewhere here? Or just in Discussion?
We compared sc2ts against three state-of-the-art methods that detect inter-lineage recombinants:
(1) RecombinHunt \cite{Alfonsi2024},
(2) CovRecomb \cite{Li2024}, and
(3) rebar (https://github.com/phac-nml/rebar).
These non-phylogenetic methods use the same underlying approach: 
they scan candidate recombinant sequences for mutations associated with different Pango lineages.

@@

\begin{table}
\begin{tabular}{llll}
 \hline
 \textbf{Method} & \textbf{Concordant in parental lineages (\%)} & \textbf{Concordant in breakpoint interval (\%)} & \textbf{Distance (bp)} \\ [1ex]
 \hline
 Sc2ts & XX & XX & XX \\ [1ex]
 \hline
 RecombinHunt-GISAID & XX & XX & XX \\ [1ex]
 \hline
 RecombinHunt-Nextstrain & XX & XX & XX \\ [1ex]
 \hline
 CovRecomb & XX & XX & XX \\ [1ex]
 \hline
 Rebar & XX & XX & XX \\ [1ex]
 \hline
\end{tabular}
\end{table}

@@

We find an overall preponderance of breakpoint intervals towards the 3’ end of the genome,
particularly at the left boundary of the Spike gene (\ref{fig:recombinant_intervals}),
consistent with the pattern of excess observed in previous studies \cite{Turakhia2022,Li2024}.
While this might result from a higher recombination rate in this region,
it may also be due to the higher levels of polymorphism in the Spike gene,
allowing recombination endpoints to be estimated more precisely.
Reassuringly, sites having unusually high mutation counts are absent in this region (\ref{fig:recombinant_intervals}),
which would confound recombination detection.

Furthermore, we find a negative relationship between
breakpoint interval size and the amount of divergence between recombinant parents,
which is measured by time to their MRCA (\ref{fig:recombinant_intervals}).
When the two parental lineages have diverged for longer periods of time (the rightmost points),
more information becomes available on either side of a breakpoint to pinpoint its location, resulting in shorter intervals.
For instance, the breakpoint intervals tend to be the smallest for events involving recombination
between the deeply divergent Omicron BA.1 and BA.2 sublineages (the red points in \ref{fig:recombinant_intervals}).

%%% TODO: Old version as placeholder. This figure is not yet finalised.
\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{recombinant_intervals}
    \caption{Recombination intervals. (A) Distribution of the breakpoint
intervals of the 356 robust recombination events along the genome
(annotated with gene labels on top). Each line represents an interval colored by its size.
(B) Number of overlapping breakpoint intervals.
(C) FIXME Heatmap showing the co-occurrence counts of the parents classified by Scorpio designations.
(D) Relationship between breakpoint interval size and time between the recombination event
and the parental MRCA. The black curve shows the theoretical expected relationship
between these quantities, estimated by minimizing the sum of squared residuals
given a single parameter (the mutation rate) (STAR Methods).}
    \label{fig:recombinant_intervals}
\end{figure}

@@

We contribute several novel methods and concepts which may find applications beyond sc2ts.
First, we devise a simplified LS model that extends the classic concept of parsimony.
In sc2ts, this model allows for accelerated HMM likelihood calculations,
enabling reconstruction of the pandemic-scale ARG presented here.
Existing HMMs used in statistical genetics may be modified in a similar way for significant speedup.
Second, we propose the concept of mutations averted by recombination,
which leads to an interpretable quantity to weigh the support for 
the hypothesis that a given sample is a recombinant against the hypothesis that it is a non-recombinant,
considering the sampled genetic diversity in an ARG and a specific HMM.
Third, we propose the concept of HMM cost to identify poorly explained samples
which are either too divergent or have a high number of mutations inconsistent with their collection dates.
Here we use HMM cost to reject time traveling samples,
which no existing phylogenetic method detects and handles during inference
(e.g., in the Viridian UShER phylogeny, parent and child nodes do not necessarily follow time order \cite{Hunt2024}),
resulting in a cleaner dataset without samples with wrong collection dates that may affect downstream analyses.
Future work may be estimating dates for the suspected time traveling samples using the ARG provided here
and the sc2ts sample matching engine.

Recombination is an important evolutionary force in viruses \cite{Perez-Losada2015},
but has been understudied in part due to the lack of suitable data models, methods, and tools \cite{Neches2020}.
While the methods and tools developed here are tailored for SARS-CoV-2, they can be extended and adapted for other viruses.
A major advantage of the sc2ts methods and tools is that they are built on top of the mature software libraries of tskit and VCF-Zarr,
which enable pandemic-scale analyses that can be performed on typical laptops.
Both tskit and VCF-Zarr are developed in Python and leverage the popular PyData ecosystem,
making them highly adaptable for a broad range of applications.
Together with the pandemic-scale ARG, we hope that sc2ts will be highly accessible to many researchers,
enabling and accelerating studies of SARS-CoV-2 for years to come.
